An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-12-31 simplified_sdc_intro/paper.tex
fix sign issue
some updates

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity)
as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep:
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le f_e\, \min_{i} \left\{\frac{e_{i}}{\dot{S}_{i}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le f_X\, \min_{i} \left\{\min_{k,X_k > \epsilon_X}\frac{{X_k}_{i}}{{\omegadot(X_k)_i}}\right\}
\end{align}
where $i$ is the zone index and $f_e$ and $f_X$ are runtime parameters
used to control the allowed change, and only species for which $X_k > \epsilon_X$ are considered.

@@

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
where $\Advs{\Uc} =  \frac{\partial \Fb^\xv (\Uc)}{\partial x} + \Hb(\Uc)$.
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  The compressible version for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

For this study we will use the 19 nuclei network containing
\isot{H}{1}, \isot{He}{3}, \isot{He}{4}, \isot{C}{12}, \isot{N}{14},
\isot{O}{16}, \isot{Ne}{20}, \isot{Mg}{24}, \isot{Si}{28},
\isot{S}{32}, \isot{Ar}{36}, \isot{Ca}{40}, \isot{Ti}{44},
\isot{Cr}{48}, \isot{Fe}{52}, \isot{Fe}{54}, \isot{Ni}{56}, protons
(from photodisintegration), and neutrons.  This is based on the
``aprox19'' network from \cite{aprox19} and originally described in
\cite{Kepler}.  We combine this with a table that gives the nuclear
statistical equilibrium (NSE) abundances in regions where the system
is in NSE.  The NSE table was generated using a 127 nuclei reaction
network and is the same as described in \cite{ma:2013}.  In our
simulations, we carry all 19 isotopes in the main network in each zone
and advect them in the hydrodynamics portion of the algorithm.  The
composition of the larger 127 nuclei network is mapped into the 19
isotopes we carry according to Table \ref{table:nuclei}.

@@

      The goal is to update the composition and thermodynamics due to the 
      change in the nuclei abundances over the (half-)timestep.  There are
      a few approaches we can do here.

      \begin{itemize}
      \item Method I: {\em Use the time-level $n$ information for the update}.

        This is the approach used in \citet{ma:2013}.  We call the table with the 
        input state:
        \begin{equation}
          \nse{\rho_{\rm in}, T_{\rm in}, (Y_e)_{\rm in}} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}

        We can now update $Y_e$:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and take the auxiliary state values of $\bar{A}$ and $(B/A)$ to be simply
        the values from the table, $\bar{A}_\out$ and $(B/A)_\out$, and the mass fractions
        $X_k$ to be the $(X_k)_{\rm out}$ returned from the table.

        Finally, we compute the energy generation rate, $\Sdot$ as:
        \begin{equation}
          \label{eq:nse_energy}
          \Sdot = \left [ \left ( \frac{B}{A} \right )_{\rm out} -
            \left ( \frac{B}{A} \right )_{\rm in} \right ] N_A \frac{1}{\Delta t}
        \end{equation}

        We note that \citet{ma:2013} modified Eq.~\ref{eq:nse_energy}
        to take use only a fraction (0.3) of the energy change from
        the table, to avoid a potential instability that can arise if
        too much energy is deposited in a zone in a single timestep
        (they called this an inertia term).  We prefer to deal with
        the issue of too much energy being deposited in a zone in one
        step through the retry mechanism already built into \castro.

      \item Method II: {\em predictor-corrector.}

        This starts out the same as the above method:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}^\star, (X_k)^\star, (B/A)^\star, dY_e/dt
        \end{equation}
        but we indicate with a $\star$ that most of those values are
        provisional, and we will seek a better value in the corrector step.
        We only update $Y_e$ from this call:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and then we use the table again, but with the updated $Y_e$
        (note that $\rho_{\rm out}=\rho_{\rm in}$ in the Strang reaction formulation):
        \MarginPar{Indicate what $T$ is used.}
        \begin{equation}
          \nse{\rho_{\rm out}, T_{\rm ???}, (Y_e)_\out} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}
        This corrects the composition so that it is consistent with the updated
        $Y_e$.  We use these values $\bar{A}_\out$, $(X_k)_\out$, and $(B/A)_\out$ to then complete the update, computing the energy as in Eq.~\ref{eq:nse_energy}.  \MarginPar{we could imagine looping over the EOS to get an updated T and then the table to get the updated $\bar{A}$, to see if that converges?}

@@

With SDC evolution, when we are in NSE, we need to do the advective
and reactive updates together.  With the tabulated NSE reaction
values, we get instantaneous values of the reaction sources, but the only
time derivative we have is for $Y_e$.  We will
use a basic predictor-corrector update:
\begin{equation}
\Uc^\star = \Uc^n + \Delta t \Adv{\Uc}^{n+1/2} + \Delta t \Rb (\Uc^n)
\end{equation}
Calling the NSE table on the starting state gives use:
\begin{equation}
\nse{\rho^n, T^n, (Y_e)^n} \rightarrow \bar{A}^n, (X_k)^n, (B/A)^n, (dY_e/dt)^n
\end{equation}
To get the reactive source, we need to use the iteratively-lagged information
from $\Ic_q^{n+1/2,(k-1)}$:
\begin{equation}
\Rb(\Uc^n) \approx \left (\begin{array}{c}
  0 \\
  \rho^n \Ics_q(\omegadot(X_k)) \\
  \rho^n \Ics_q(\omegadot(\alpha_l)) \\
  0 \\
  \Ics_q(\rho e) 
\end{array}
\right )
\end{equation}
Since we have $(dY_e/dt)^n$ from the table, we replace the source for $Y_e$ in the above
$\Rb(\Uc^n)$ as:
\begin{equation}
R(\rho Y_e) = \rho (dY_e/dt)^n
\end{equation}
which gives an update for the electron fraction as:
\begin{equation}
(\rho Y_e)^\star = (\rho Y_e)^n + \Delta t \Advss{\rho Y_e}^{n+1/2} + \Delta t \rho (dY_e/dt)^n
\end{equation}
From this prediction, we get $\rho^\star$, $\alpha_l^\star$, and
$e^\star$, which allows us to call the EOS and get an updated
temperature, $T^\star$. \MarginPar{Andy suggests time-centering $dY_e/dt$}
We can then again call the NSE table:
\begin{equation}
\nse{\rho^\star, T^\star, (Y_e)^\star} \rightarrow \bar{A}^{n+1,(k)}, (X_k)^{n+1,(k)}, (B/A)^{n+1,(k)}, (dY_e/dt)
\end{equation}
where we use the outputs to define the new values of the composition variables.
Since the density and momenta don't have reactive sources, we just use the $\Uc^\star$ state
as their new values:
\begin{align}
\rho^{n+1,(k)} &= \rho^\star \\
(\rho u)^{n+1,(k)} &= (\rho u)^\star
\end{align}
We then compute the energy release as
\begin{equation}
       (\rho \Sdot)^{n+1/2} = \left [ \rho^{n+1,(k)} \left ( \frac{B}{A} \right )^{n+1,(k)} -
                            \rho^n \left ( \frac{B}{A} \right )^n \right ] N_A \frac{1}{\Delta t}
\end{equation}\MarginPar{is this right?}
and then do the final update of the total energy:
\begin{equation}
(\rho E)^{n+1,(k)} = (\rho E)^n + \Delta t \Advss{\rho E}^{n+1/2} + (\rho \Sdot)^{n+1/2}
\end{equation}

>>>>

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity)
as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep: \MarginPar{do we enforce this in the NSE region?}
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le f_e\, \min_{i} \left\{\frac{e_{i}}{\dot{S}_{i}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le f_X\, \min_{i} \left\{\min_{k,X_k > \epsilon_X}\frac{{X_k}_{i}}{{\omegadot(X_k)_i}}\right\}
\end{align}
where $i$ is the zone index and $f_e$ and $f_X$ are runtime parameters
used to control the allowed change, and only species for which $X_k > \epsilon_X$ are considered.

@@

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
where $\Advs{\Uc} =  -{\partial \Fb^\xv (\Uc)}/{\partial x} + \Hb(\Uc)$.
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  The compressible version for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

For this study we will use the 19 nuclei network containing
\isot{H}{1}, \isot{He}{3}, \isot{He}{4}, \isot{C}{12}, \isot{N}{14},
\isot{O}{16}, \isot{Ne}{20}, \isot{Mg}{24}, \isot{Si}{28},
\isot{S}{32}, \isot{Ar}{36}, \isot{Ca}{40}, \isot{Ti}{44},
\isot{Cr}{48}, \isot{Fe}{52}, \isot{Fe}{54}, \isot{Ni}{56}, protons
(from photodisintegration), and neutrons.  This is based on the
``aprox19'' network from \cite{aprox19} and originally described in
\cite{Kepler}.  We use a modified version of the VODE~\citep{vode} integrator---ported to C++ with checks added to the timestep rejection logic that ensure that the species mass fractions stay between 0 and 1\footnote{This modified version of VODE is available in our Microphysics repo: \url{https://github.com/starkiller-astro/Microphysics}.}.\MarginPar{discuss tolerances}  We combine this with a table that gives the nuclear
statistical equilibrium (NSE) abundances in regions where the system
is in NSE.  The NSE table was generated using a 127 nuclei reaction
network and is the same as described in \cite{ma:2013}.  In our
simulations, we carry all 19 isotopes in the main network in each zone
and advect them in the hydrodynamics portion of the algorithm.  The
composition of the larger 127 nuclei network is mapped into the 19
isotopes we carry according to Table \ref{table:nuclei}.

@@

      The goal is to update the composition and thermodynamics due to the 
      change in the nuclei abundances over the (half-)timestep.  \citet{ma:2013}
      uses a first-order in time difference to get the new composition state
      and evaluates the energy release from the change in binding energy in the NSE
      state.  They only apply a fraction (0.3) of the energy release to the internal
      energy in a zone, to avoid a potential instability that can arise if too much
      energy is added to a zone in a single timestep.  We prefer to deal with this
      issue through the retry mechanism already built into \castro.

      The approach we use begins with calling the NSE table with the input state:
      \begin{equation}
        \nse{\rho_\inp, T_\inp, (Y_e)_\inp} \rightarrow \bar{A}^\star, (X_k)^\star, (B/A)^\star, dY_e/dt
        \end{equation}
      We indicate with a $\star$ that most of those values are
      provisional, and we will seek a better value in the corrector step.
      We only update $Y_e$ from this call:
      \begin{equation}
        (Y_e)_\out = (Y_e)_\inp + \Delta t \frac{dY_e}{dt}
      \end{equation}
      and then we use the table again, but with the updated $Y_e$
      (note that $\rho_{\rm out}=\rho_{\rm in}$ in the Strang reaction formulation):
      \begin{equation}
        \nse{\rho_\out, T_\inp, (Y_e)_\out} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
      \end{equation}
      In this formulation, we have not updated the temperature. \MarginPar{discuss this more, and perhaps switch to a mode where we try to predict T}
      This corrects the composition so that it is consistent with the updated
      $Y_e$.  We use these values $\bar{A}_\out$, $(X_k)_\out$, and $(B/A)_\out$ to then complete the update, computing the energy release, $\Sdot$ as:
      \begin{equation}
        \label{eq:nse_energy}
        \Sdot = \left [ \left ( \frac{B}{A} \right )_\out -
          \left ( \frac{B}{A} \right )_\inp \right ] N_A \frac{1}{\Delta t}
      \end{equation}
      \MarginPar{we could imagine looping over the EOS to get an updated T and then the table to get the updated $\bar{A}$, to see if that converges?}

@@

With SDC evolution, when we are in NSE, we need to do the advective
and reactive updates together.  With the tabulated NSE reaction
values, we get instantaneous values of the reaction sources, but the only
time derivative we have is for $Y_e$.  We will
use a basic predictor-corrector update:
\begin{equation}
\Uc^\star = \Uc^n + \Delta t \Adv{\Uc}^{n+1/2} + \Delta t \left [\Rb (\Uc) \right ]^{n+1/2}
\end{equation}
Calling the NSE table on the starting state gives us:
\begin{equation}
\nse{\rho^n, T^n, (Y_e)^n} \rightarrow \bar{A}^n, (X_k)^n, (B/A)^n, (dY_e/dt)^n
\end{equation}
To get the reactive source, we need to use the iteratively-lagged time-centered information
from $\Ic_q^{n+1/2,(k-1)}$.  We only need to update the internal energy and auxiliary data,
since together with the density, that is all that is needed for the equation of state to get the
updated temperature.    So our sources are:
\begin{equation}
\Rb(\rho e) \approx \Ics_q(\rho e)
\end{equation}
and
\begin{equation}
\Rb(\rho \alpha_l) \approx \rho^{n+1/2} \Ics_q(\alpha_l)
\end{equation}
were $\rho^{n+1/2} = (\rho^n + \rho^\star)/2$.

Since we have $(dY_e/dt)^n$ from the table, we replace the source for $Y_e$ in the above
$\Rb(\Uc^n)$ as: \MarginPar{it might be better if we had $\Ics_\Uc$}
\begin{equation}
R(\rho Y_e) = \rho^{n+1/2} (dY_e/dt)^n
\end{equation}
which gives an update for the electron fraction as:
\begin{equation}
(\rho Y_e)^\star = (\rho Y_e)^n + \Delta t \Advss{\rho Y_e}^{n+1/2} + \Delta t \rho^{n+1/2} (dY_e/dt)^n
\end{equation}
From this prediction, we get $\rho^\star$, $\alpha_l^\star$, and
$e^\star$, which allows us to call the EOS and get an updated
temperature, $T^\star$. \MarginPar{Andy suggests time-centering $dY_e/dt$}
We can then again call the NSE table:
\begin{equation}
\nse{\rho^\star, T^\star, (Y_e)^\star} \rightarrow \bar{A}^{n+1,(k)}, (X_k)^{n+1,(k)}, (B/A)^{n+1,(k)}, (dY_e/dt)
\end{equation}
where we use the outputs to define the new values of the composition variables.
Since the density and momenta don't have reactive sources, we just use the $\Uc^\star$ state
as their new values:
\begin{align}
\rho^{n+1,(k)} &= \rho^\star \\
(\rho u)^{n+1,(k)} &= (\rho u)^\star
\end{align}
We then compute the energy release as
\begin{equation}
       (\rho \Sdot)^{n+1/2} = \left [ \rho^{n+1,(k)} \left ( \frac{B}{A} \right )^{n+1,(k)} -
                            \rho^n \left ( \frac{B}{A} \right )^n \right ] N_A \frac{1}{\Delta t}
\end{equation}\MarginPar{is this right?}
and then do the final update of the total energy:
\begin{equation}
(\rho E)^{n+1,(k)} = (\rho E)^n + \Delta t \Advss{\rho E}^{n+1/2} + (\rho \Sdot)^{n+1/2}
\end{equation}
